U.S. Labour Market Resilience

### 2024-03-14 outputs/paper/paper.qmd
update paper

---
title: "Labour Market Resilience"
subtitle: "Economic Crisis Has Diverse Impacts on Industry Work Hours"
author: 
  - Mingjia Chen
  - Catherine Punnoose
  - Tianen (Evan) Hao
thanks: "Code and data are available at: https://github.com/ShadyEvan4830/Labour_Market_Resilience.git"
date: March 11, 2024
date-format: long
abstract: "Historically, economic crises have impacted labour market working patterns and employment levels, but are all groups within the labour market suffering from overall similar trends? This study uses data from the U.S. General Social Survey (GSS) to investigate the impact of major economic downturns on U.S. labour force dynamics, focusing specifically on “hours worked last week.” Analyzing changes in working hours reveals significant reductions, particularly affecting younger workers, those with less education, and women. These findings are important since they underscore the necessity for targeted interventions to support the most vulnerable groups in the labour market."
format: pdf
toc: true
number-sections: true
bibliography: bibliography.bib
---

@@

# Introduction

Economic downturns from the 1998 Asian Financial Crisis and 2008 Global Financial Crisis to the unprecedented 2020 COVID-19 pandemic, have significantly shaped labour market landscapes, impacting employment patterns and work hours across the United States. These pivotal events, each unique in their severity and global reach, provide a crucial backdrop for examining the labour market's resilience and adaptability amidst fluctuating economic conditions. Yet, a critical question arises: How uniformly do these economic shifts affect various segments within the labour market, particularly regarding working hours?

To explore this question, this paper focuses on the U.S. General Social Survey (GSS) [@GSS] data to concentrate on the "number of hours worked last week" as a primary metric for labour engagement. This study is predicated on the hypothesis that economic declines precipitate a decrease in labour engagement, evidenced by reduced work hours, with varying degrees of impact across different labour groups. Through a comparative analysis of periods marked by economic crises against times of economic stability, we aim to dissect the labour market's response mechanisms, spotlighting the sectors and demographic groups most susceptible to economic adversities.

By analyzing the comprehensive GSS data, including variables such as gender, age, and career field, we seek to offer a granular understanding of how major economic crises influence work hours, thereby contributing to a broader discourse on labour market adaptability in the face of economic instability.

Initial analysis reveals notable disparities: younger workers, those with less education, and women experience significant reductions in work hours during economic crises. This finding highlights the differential impact of economic downturns across demographic groups and signals the urgent need for targeted policy interventions to mitigate these disparities and bolster labour market resilience.

The ensuing sections of this paper will outline the data's nature and scope, detail the analytical process and findings, and discuss these findings' broader implications for labour market dynamics amidst economic crises. The conclusion will synthesize the key insights derived from this analysis, and an appended supplementary survey will propose additional inquiries to explore labour market responses to economic downturns further.

@@

@@

ggplot(data, aes(x = "", y = X1998, fill = work_hours)) +
  geom_bar(stat = "identity", width = 1) +
  coord_polar("y", start = 0) +
  scale_fill_brewer(palette = "Set3") +  # Use a color palette
  theme_void() +  # Remove background and gridlines
  labs(title = "1998 Survey Results") +
  theme(plot.title = element_text(hjust = 0.5, size = 16))
```
\break

@@

ggplot(data, aes(x = "", y = X2008, fill = work_hours)) +
  geom_bar(stat = "identity", width = 1) +
  coord_polar("y", start = 0) +
  scale_fill_brewer(palette = "Set3") +  # Use a color palette
  theme_void() +  # Remove background and gridlines
  labs(title = "2008 Survey Results") +
  theme(plot.title = element_text(hjust = 0.5, size = 16))
```
\break

@@

ggplot(data, aes(x = "", y = X2021, fill = work_hours)) +
  geom_bar(stat = "identity", width = 1) +
  coord_polar("y", start = 0) +
  scale_fill_brewer(palette = "Set3") +  # Use a color palette
  theme_void() +  # Remove background and gridlines
  labs(title = "2021 Survey Results") +
  theme(plot.title = element_text(hjust = 0.5, size = 16))
```

@@

# Results

@@

# Discussion

@@

@@

# Conclusion

>>>>

---
title: "U.S. Labour Market Resilience"
subtitle: "Impact Analysis of Economic Crises on Industry Work Hours in 1998, 2008 and 2020"
author: 
  - Mingjia Chen
  - Catherine Punnoose
  - Tianen (Evan) Hao
thanks: "Code and data are available at: https://github.com/ShadyEvan4830/Labour_Market_Resilience.git"
date: March 11, 2024
date-format: long
abstract: "Economic crises have impacted labour market working patterns and employment levels, but are all groups within the labour market suffering from similar trends? Focusing on the Asian Financial Crisis, the Global Financial Crisis, and the COVID-19 pandemic, this study uses data from the U.S. General Social Survey (GSS) to investigate the impact of major economic downturns on U.S. labour force dynamics, focusing specifically on “hours worked last week.” Our analysis reveals that economic downturns lead to a decline in average working hours, and identifies disparities in the effects of economic crises on demographic groups, with women and younger workers disproportionately affected. These findings underscore the necessity for targeted interventions to support the most vulnerable groups in the labour market."
format: pdf
toc: true
number-sections: true
bibliography: bibliography.bib
---

@@

# Introduction {#sec-intro}

Economic downturns, from the 1998 Asian Financial Crisis and 2008 Global Financial Crisis to the 2020 COVID-19 pandemic, have significantly impacted the U.S. labour market, affecting employment patterns and work hours. These events, unique in severity and reach, provide a crucial backdrop for examining labour market resilience and adaptability. A critical question arises: How uniformly do these shifts affect labour market segments, especially in work hours?

This paper uses U.S. General Social Survey (GSS) [@GSS] data to examine the "number of hours worked last week" as a key measure of labour engagement. We hypothesize that economic downturns lead to decreased labour engagement, seen in reduced work hours, impacting labour groups differently. By comparing periods of economic crises with times of stability, we aim to understand the labour market's responses, focusing on the most affected sectors and demographic groups.

Our analysis of GSS data, including gender, age, and career field variables, seeks to provide detailed insights into how economic crises affect work hours, contributing to discussions on labour market flexibility during economic instability.

Initial findings show disparities: younger workers, those with less education, and women face significant work-hour reductions during crises. This underscores the need for targeted policy interventions to address these disparities and strengthen labour market resilience.

This paper begins with an [Introduction] framing the impact of economic downturns on work hours. It is followed by the [Data] section detailing methodology and cleaning efforts. In the [Results], we analyze work hours and nonresponse rates, leading to the [Discussion] on labour market responses to crises, including the impact of COVID-19, and addressing ethical issues. We later acknowledge our paper's Limitations, suggest paths for Future Research, especially on enhancing response rates, and conclude with a synthesis of our insights. The [Appendix] provides further survey information.

@@

From the dataset, we used the variable “response” to collect the respondent’s responses. For clarity, the variable was renamed to “work hours” as the respondent’s response was the number of hours they worked. Another variable used is “year”. The purpose of this variable is to collect the year the survey was conducted.

@@

```
\break

@@

```
\break

@@

@@

# Results {#sec-results}

@@

# Discussion {#sec-discussion}

@@

##  Policy Interventions: Short-Term Relief and Long-Term Resilience

The significant drop in work hours during economic downturns underscores the importance of policy interventions and social welfare to support both immediate recovery and long-term labour market resilience. For example, during the COVID-19 pandemic, expanded unemployment benefits and direct financial aid provided critical support demonstrating effective short-term relief strategies, as @MacKinnon mentioned in his article.

Furthermore, long-term measures, such as retraining programs for emerging sectors, emphasize the need for policies that bolster workforce adaptability. By equipping workers with new skills, these programs facilitate transitions to stable employment in growing industries. For the gig economy, particularly impacted by economic crises, requires policies ensuring social security benefits and income stability. Implementing health and safety standards for gig workers can also enhance their job security and well-being.

@@

# Conclusion {#sec-conclusion}
